---
title: "Fenton & Tureson 573 Final Project"
author: "Laura Fenton & Kayla Tureson"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-pkg, message = FALSE}
library(ggplot2)  # for plots
library(tinytex)
library(here)
library(readxl)  # for reading excel files
library(modelsummary)  # for summarizing data
library(rstan)
rstan_options(auto_write = TRUE)  # save compiled STAN object
options(mc.cores = 2)  # use two cores
library(posterior)
library(bayesplot)
library(dplyr)
library(kableExtra)
library(dagitty)
library(brms)
library(ggdag)  # render DAGs in ggplot style
theme_set(theme_classic() +
    theme(panel.grid.major.y = element_line(color = "grey92")))
```

# Research Questions

> Is there a difference in amyloid accumulation between individuals in the control condition and the exercise intervention condition?
> Is there a difference in hippocampal atrophy between individuals in the control condition and the exercise intervention condition?
> Does VO2 max change mediate the relationship between treatment condition and hippocampal atrophy rate?

# Variables

- `amyloid_accum`: accumulation of amyloid from pre (baseline) to post (wk52) intervention
- `Hippocampus_change`: hippocampal atrophy from (baseline) to post (wk52) intervention
- `VO2_change`: change in VO2 max from (baseline) to post (wk52) intervention
- `apx_group`: AExSupport = aerobic exercise condition, NoSupport = control condition

## Import Data

```{r exercise}
# load the data
exercise <- read_excel(here("APEx_FinalData_20201021.xlsx"))

# subset only individuals who have baseline and follow up data
tt <- table(exercise$study_id)
exercise2 <- subset(exercise, study_id %in% names(tt[tt == 3]))

# log transform key variables
exercise2$amyloid_Mean6_CB <- log(exercise2$amyloid_Mean6_CB)
exercise2$Hippocampus_Vol_mL <- log(exercise2$Hippocampus_Vol_mL)
exercise2$apx_exppt_vo2mx_ml <- log(exercise2$apx_exppt_vo2mx_ml)

# create new variables: (amyloid_accum) for amyloid accumulation from baseline 
# to week 52 (smaller values are better); (Hippocampus_change) for hippocampal neurodegeneration from baseline to week 52 (smaller values are better); (VO2_change) for increase in VO2 max from baseline to 52 (larger values are better)
exercise3 <- 
exercise2 %>% 
  group_by(study_id)  %>% 
  mutate(amyloid_accum = amyloid_Mean6_CB[timept=="WK52"] -
           amyloid_Mean6_CB[timept=="BL"],
         Hippocampus_change = Hippocampus_Vol_mL[timept=="BL"] -
           Hippocampus_Vol_mL[timept=="WK52"],
         VO2_change = apx_exppt_vo2mx_ml[timept=="WK52"] -
           apx_exppt_vo2mx_ml[timept=="BL"])

# subset only the first entry of each participant 
 exercise.subset <- 
  exercise3 %>% 
  group_by(study_id) %>% 
  filter(row_number()==1)

# subset only the third entry of each participant 
exercise.subset.wk52 <- 
  exercise3 %>% 
  group_by(study_id) %>% 
  filter(row_number()==3)

# change the condition variable to a factor
exercise.subset$apx_group <- as.factor(exercise.subset$apx_group)

# remove NAs so that STAN will run later
exercise.subset <- na.omit(exercise.subset)

# create new variable for condition and set as integer
exercise.subset$apx_group_coded <- ifelse(exercise.subset$apx_group=="NoSupport", 0, ifelse(exercise.subset$apx_group=="AExSupport", 1, NA))

exercise.subset$apx_group_coded <- as.integer(exercise.subset$apx_group_coded)
```

## Visualize data

```{r visualize}
# Visualize baseline variables
ggplot(exercise.subset, aes(x=amyloid_Mean6_CB, color=apx_group)) +
  geom_histogram(fill="white", alpha=0.5, position="identity")

ggplot(exercise.subset, aes(x=Hippocampus_Vol_mL, color=apx_group)) +
  geom_histogram(fill="white", alpha=0.5, position="identity")

ggplot(exercise.subset, aes(x=apx_exppt_vo2mx_ml, color=apx_group)) +
  geom_histogram(fill="white", alpha=0.5, position="identity")

#Visualize week 52 variables

ggplot(exercise.subset.wk52, aes(x=amyloid_Mean6_CB, color=apx_group)) +
  geom_histogram(fill="white", alpha=0.5, position="identity")

ggplot(exercise.subset.wk52, aes(x=Hippocampus_Vol_mL, color=apx_group)) +
  geom_histogram(fill="white", alpha=0.5, position="identity")

ggplot(exercise.subset.wk52, aes(x=apx_exppt_vo2mx_ml, color=apx_group)) +
  geom_histogram(fill="white", alpha=0.5, position="identity")

# Visualize change variables
ggplot(exercise.subset, aes(x=amyloid_accum, color=apx_group)) +
  geom_histogram(fill="white", alpha=0.5, position="identity")

ggplot(exercise.subset, aes(x=Hippocampus_change, color=apx_group)) +
  geom_histogram(fill="white", alpha=0.5, position="identity")

ggplot(exercise.subset, aes(x=VO2_change, color=apx_group)) +
  geom_histogram(fill="white", alpha=0.5, position="identity")
```

## Variable Summary

```{r summ-var}
## can't get N to work, not sure why 
datasummary(amyloid_accum * 
                (Mean + SD + Min + Max + Histogram) ~ 
                factor(apx_group),
            data = exercise.subset,
caption = "Summary Statistics") %>%
    # add table note
    add_footnote("SD = standard deviation", notation = "none")

datasummary(Hippocampus_change * 
                (Mean + SD + Min + Max + Histogram) ~ 
                factor(apx_group),
            data = exercise.subset,
caption = "Summary Statistics") %>%
    # add table note
    add_footnote("SD = standard deviation", notation = "none")

```

# Model Amyloid Accumulation

Let $Y_A$ = change in amyloid from baseline to week 52 (week 52 - baseline; smaller values indicate less accumulation), $G$ = treatment condition
Model:
$$
  \begin{aligned}
    Y_{Ai, G = 0} & \sim N(\mu_1, \sigma_1) \\
    Y_{Ai, G = 1} & \sim N(\mu_2, \sigma_2)
  \end{aligned}
$$

Prior:
$$
  \begin{aligned}
    \mu_1 & \sim N(3, 2) \\
    \mu_2 & \sim N(3, 2) \\
    \sigma_1 & \sim N^+(0, 2) \\
    \sigma_2 & \sim N^+(0, 2)
  \end{aligned}
$$

## Running Stan

We used 4 chains, each with 4,000 iterations (first 2,000 as warm-ups). 

```{r m1, results = "hide"}
# Rescale data so all are greater than 0 and STAN will run (chose .13 based 
# on min values from summary data)
exercise.subset$amyloid_accum_rescale <- exercise.subset$amyloid_accum + .13

# 1. form the data list for Stan
stan_dat <- with(exercise.subset,
    list(N1 = sum(apx_group == "NoSupport"),
         N2 = sum(apx_group == "AExSupport"),
         y1 = amyloid_accum_rescale[which(apx_group == "NoSupport")],
         y2 = amyloid_accum_rescale[which(apx_group == "AExSupport")])
)

# 2. Run Stan
m1 <- stan(
    file = here("robust_2group.stan"),
    data = stan_dat,
    seed = 1234,  # for reproducibility
    iter = 4000
)
```

# Results

As shown in the graph below, the chains mixed well.

```{r rank-hist-m1}
mcmc_rank_hist(m1, pars = c("mu1", "mu2", "sigma1", "sigma2"))
```

The following table shows the posterior distributions of $\mu_1$, $\mu_2$, $\sigma_1$, $\sigma_2$, and $\mu_2 - \mu_1$.

```{r summ-m1}
summ_m1 <- as_draws_df(m1) %>%
    subset_draws(variable = c("mu1", "mu2", "sigma1", "sigma2")) %>%
    mutate_variables(`mu2 - mu1` = mu2 - mu1) %>%
    summarise_draws()
knitr::kable(summ_m1, digits = 2)
```

The analysis showed that on average, there was no difference in amyloid accumulation from baseline to week 52 between individuals in the aerobic exercise condition and individuals in the education control condition. Individuals in both the control condition and intervention condition had a posterior mean of 0.14. The 90% CI's were identical for both groups. As a reminder, the actual values are .14 lower than those we see in the table, due to data transformation prior to running our model.

# Model Hippocampal Neurodegeneration

Let $Y_A$ = change in amyloid from baseline to week 52 (week 52 - baseline; smaller values indicate less accumulation), $G$ = treatment condition
Model:
$$
  \begin{aligned}
    Y_{Hi, G = 0} & \sim N(\mu_1, \sigma_1) \\
    Y_{Hi, G = 1} & \sim N(\mu_2, \sigma_2)
  \end{aligned}
$$

Prior:
$$
  \begin{aligned}
    \mu_1 & \sim N(3, 2) \\
    \mu_2 & \sim N(3, 2) \\
    \sigma_1 & \sim N^+(0, 2) \\
    \sigma_2 & \sim N^+(0, 2)
  \end{aligned}
$$

## Running Stan

We used 4 chains, each with 4,000 iterations (first 2,000 as warm-ups). 

```{r m2, results = "hide"}
# Rescale data so all are greater than 0 and STAN will run (chose .03 based 
# on min values from summary data)
exercise.subset$Hippocampus_change_rescale <- exercise.subset$Hippocampus_change + .03

# 1. form the data list for Stan
stan_dat_2 <- with(exercise.subset,
    list(N1 = sum(apx_group == "NoSupport"),
         N2 = sum(apx_group == "AExSupport"),
         y1 = Hippocampus_change_rescale[which(apx_group == "NoSupport")],
         y2 = Hippocampus_change_rescale[which(apx_group == "AExSupport")])
)
# 2. Run Stan
m2 <- stan(
    file = here("robust_2group.stan"),
    data = stan_dat_2,
    seed = 1234,  # for reproducibility
    iter = 4000
)
```

# Results

As shown in the graph below, the chains mixed well.

```{r rank-hist-m2}
mcmc_rank_hist(m2, pars = c("mu1", "mu2", "sigma1", "sigma2"))
```

The following table shows the posterior distributions of $\mu_1$, $\mu_2$, $\sigma_1$, $\sigma_2$, and $\mu_2 - \mu_1$.

```{r summ-m2}
summ_m2 <- as_draws_df(m2) %>%
    subset_draws(variable = c("mu1", "mu2", "sigma1", "sigma2")) %>%
    mutate_variables(`mu2 - mu1` = mu2 - mu1) %>%
    summarise_draws()
knitr::kable(summ_m2, digits = 2)
```

The analysis showed that on average, there was no difference in hippocampal neurodegeneration from baseline to week 52 between individuals in the aerobic exercise condition and individuals in the education control condition. Individuals in the exercise and control conditions both had a posterior mean of 0.04, and the confidence intervals were identical. As a reminder, the actual values are .03 lower than those we see in the table, due to data transformation prior to running our model.

## Mediation

```{r dag1}
dag1 <- dagitty(
    "dag{
      C -> HV; C -> V; V -> HV
    }"
)
coordinates(dag1) <- list(x = c(C = 0, V = 1, HV = 2),
                          y = c(C = 0, V = 1, HV = 0))
# Plot
ggdag(dag1) + theme_dag()
```
```{r}
get_prior(Hippocampus_change ~ apx_group_coded + VO2_change,
          data = exercise.subset,
          family = gaussian(link = "identity"))

get_prior(VO2_change ~ apx_group_coded,
         data = exercise.subset,
         family = gaussian(link = "identity"))

prior_m <- prior(normal(0, 1), class = "b") 
m_med <- brm(
        bf(Hippocampus_change ~ apx_group_coded + VO2_change) +
        bf(VO2_change ~ apx_group_coded) +
        set_rescor(FALSE),
       family = list(gaussian("identity"), gaussian("identity")),
       data = exercise.subset,
       prior = prior(normal(1,1), class = "b", resp = "VO2change") +
        prior(student_t(3, 0, 2.5), class = "sigma", resp = "VO2change") +
        prior(normal(1, 1), class = "b", resp = "Hippocampuschange"),
      seed = 1338,
      iter = 4000
)
print(m_med)
```



$$VO2 change_{i} \sim N(\mu^v_{i}, \sigma^jv)$$
$$\mu^v_{i} = \beta^v_0 + \beta_1group_i$$
$$hippocampuschange_{i} \sim N(\mu^h_{i}, \sigma^h)$$
$$\mu^h_{i} = \beta^h_0 + \beta_2VO2change_i +\beta_3group_{i}$$
$$\beta^v_0, \beta^h_0 \sim N(0, 1)$$
$$\beta_1, \beta_2, \beta_3 \sim N(0,1)$$
$$\sigma \sim t^+_{3}(0,2.5)$$
```

```{r pp checks}
pp_check(m_med, resp = "VO2change")

pp_check(m_med, resp = "Hippocampuschange")

pp_check(m_med, type = "error_scatter_avg_vs_x", resp = "Hippocampuschange",
x = "VO2_change")
```

